---
title: "Exact Run-Count Distributions"
author: "Bahadır Yüzbaşı"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exact Run-Count Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Model and assumptions

Let $X_1,\ldots,X_n$ denote a non-empty ordered sequence of mutually
independent Bernoulli variables. At position $i$, the supplied number
$p_i=P(X_i=1)$ is finite and belongs to $[0,1]$. The probabilities may differ
across positions and are treated as fixed inputs; the package does not
estimate them.

A run is a maximal consecutive block of equal values. Consequently, the total
run count $R_n$ is one for the first block plus one for every change between
adjacent observations:

\[
R_n=1+\sum_{i=2}^{n}\mathbf{1}(X_i\ne X_{i-1}).
\]

Here $\mathbf{1}(\cdot)$ equals one when its statement is true and zero
otherwise. This expression defines the package's random variable and is valid
for non-empty binary sequences.

## Exact distribution

For a vector of position-specific probabilities, `runs_distribution()`
returns every possible value of $R_n$, its exact probability mass, and its
inclusive lower cumulative probability.

```{r distribution}
library(BernoulliRuns)

prob <- c(0.15, 0.40, 0.75, 0.60)
distribution <- runs_distribution(prob)
knitr::kable(
  distribution,
  digits = 6,
  caption = "Exact distribution of the total run count"
)
```

The computation tracks the probability of each current run count separately
for sequences ending in zero and one. At the next position, retaining the last
value continues a run, whereas changing it opens a new run. Mutual
independence justifies multiplying each previous state probability by $p_i$
or $1-p_i$. The procedure uses quadratic time and linear working memory in
the sequence length $n$.

The expectation can be calculated without first forming the full
distribution. It is one plus the sum, over adjacent positions, of the two
possible change probabilities:

\[
E(R_n)=1+\sum_{i=2}^{n}
\{p_{i-1}(1-p_i)+(1-p_{i-1})p_i\}.
\]

The two terms inside braces are respectively the probabilities of a change
from one to zero and from zero to one. The formula relies on the same mutual-
independence and fixed-probability assumptions.

```{r expectation}
runs_mean(prob)
sum(distribution$runs * distribution$probability)
```

## Observed sequence and one-sided tails

For an observed binary sequence, `count_runs()` evaluates the same run-count
definition. `runs_tail()` then calculates an inclusive lower probability
$P(R_n\le r_{obs})$ or inclusive upper probability $P(R_n\ge r_{obs})$, where
$r_{obs}$ denotes the observed number of runs.

```{r tails}
x <- c(0, 1, 1, 0)
count_runs(x)
runs_tail(x, prob, tail = "lower")
runs_tail(x, prob, tail = "upper")
```

These are conditional probabilities under the supplied model. If the same
observations were used to estimate `prob`, a tail probability is not
automatically a calibrated exact test. The package intentionally does not
combine the two tails into a two-sided p-value and does not model dependent or
Markov sequences.

